
---
title: "MCAR 30% missing - DPMPM"
output:
  pdf_document: default
  html_document:
    highlight: pygments
    theme: spacelab
---

```{r setup, echo =FALSE, include=FALSE}
# DO NOT ALTER CODE IN THIS CHUNK
# The very first time you use this R markdown file, you should install each of the packages below.
# The same goes for other packages you might decide to use
# Remember that you only need to install each R package once in R (ever -- unless you change computers).
# All you need to do whenever you need to use the package again (after restarting the R session),
# is to use the library function to call the package.
# For example, type install.packages("knitr") in the console to install the knitr package. 
knitr::opts_chunk$set(echo = TRUE, fig.pos = 'h', fig.align = 'center')
knitr::opts_chunk$set(fig.cap = "",  fig.path = "Plot")
library(knitr)
library(dplyr)
library(arm)
library(pROC)
library(tidyverse)
library(MASS)
library(tigerstats)
library(leaps)
library(car)
library(rms)
require(caret)
require(e1071)
library(lme4) 
library(lattice)
library(broom)
library(boot)
library(ggplot2)
library(cobalt)
require(tidyverse)
require(rstanarm)
require(magrittr)
require(rstan)
require(MCMCpack)
library(abind)
library(matrixStats)
library(truncnorm)
library(mvtnorm)
library(MCMCpack)
library(mnormt)
library(coda)
library(NPBayesImputeCat)
```

* * *
```{r}
# sample MCAR dataset from PUMS
source("../../utils/sampleMCAR.R")
n = 10000
missing_col = c(1,3,7,9,10,11)
missing_prob = 0.3
set.seed(0)

output_list <- sampleMCAR(n, missing_prob)
df <- output_list[['df']]
df_observed <- output_list[['df_observed']]
```


### DPMPM

Multiple imputation using NPBayesImputeCat package

Ref: https://cran.r-project.org/web/packages/NPBayesImputeCat/NPBayesImputeCat.pdf

1. Create and initialize the Rcpp_Lcm model object using CreateModel with the following arguments:
  - X: dataframe to be imptuted = df
  - MCZ: dataframe with the definition of structural zero = NULL
  - K: the maximum number of mixture components = 40
  - Nmax: An upper truncation limit for the augmented sample size = 0
  - aalpha: the hyper parameter alpha in stick-breaking prior = 0.25
  - balpha: the hyper parameter beta in stick-breaking prior = 0.25
  - seed = 0

2. Set the tracer for the sampling process
  - k_star: the effective cluster number
  - psi: conditional multinomial probabilties
  - ImputedX: imputation result

3. Run the model using the method Run of Rcpp_Lcm class with the following arguments:
  - burnin = 10000
  - iter = 10000
  - thinning = 5

4. Obtain result

```{r, results = 'hide'}
N = 40
Mon = 10000
B = 10000
thin.int = 5

# 1. Create and initialize the Rcpp_Lcm model object
model = CreateModel(X = df_observed, MCZ = NULL, K = N, Nmax = 0,
                    aalpha = 0.25, balpha = 0.25, seed = 0)
# 2. Set tracer
model$SetTrace(c('k_star', 'psi', 'ImputedX', 'alpha'),Mon)

# 3. Run model using Run(burnin, iter, thinning)
model$Run(B,Mon,thin.int)
```


```{r}
# Extract results and format output
output <- model$GetTrace()
k_star <- output$k_star
psi <- output$psi
imputed_df <- output$ImputedX
alpha <- output$alpha

#retrieve parameters from the final iteration 
result <- model$snapshot

#convert ImputedX matrix to dataframe, using proper factors/names etc. 
ImputedX <- GetDataFrame(result$ImputedX,df)

# extract 5 imputed dataset from DP model
imputation_index = as.integer(seq(1,dim(imputed_df)[1], length.out = 5))
imputation_list = list()
levels = c(7,7,7,19,5,4,7,2,17,3,13)
for (i in 1:length(imputation_index)) {
  index = imputation_index[i]
  # need to plus 1 here because the class index of DP function starts at 0
  d = imputed_df[index,] + 1
  dim(d) = dim(t(df_observed))
  d = data.frame(t(d))
  colnames(d) = colnames(df_observed)
  # format columns of d
  for (col_index in 1:ncol(df_observed)) {
    d[,col_index] = factor(d[,col_index], levels = 1:levels[col_index], ordered = TRUE)
  }
  imputation_list[[i]] = d
}
```

Diagnostics

```{r, echo = FALSE}
# The number of clusters/latent groups used
plot(1:length(k_star), k_star, xlab = 'trials', 
     ylab = 'number of active clusters', 
     main = 'Number of clusters used over time', ylim = c(0,40))
```

```{r, echo = FALSE}
# sampled alpha value for the stick breaking process
plot(1:length(alpha), alpha, xlab = 'trials', 
     ylab = 'alpha', type = 'l',
     main = 'alpha value for the stick breaking process')
```

```{r, echo = FALSE}
alpha.mcmc <- mcmc(alpha, start=1)
# auto correlation
autocorr.plot(alpha.mcmc)
```


```{r}
source("../../utils/create_report.R")
create_report(imputation_list, max_nway=4, missing_col, df_observed)
```
* * *



