
---
title: "Dirichlet process mixture of products of multinomial distributions model (DPMPM)"
output:
  pdf_document: default
  html_document:
    highlight: pygments
    theme: spacelab
---

```{r setup, echo =FALSE, include=FALSE}
# DO NOT ALTER CODE IN THIS CHUNK
# The very first time you use this R markdown file, you should install each of the packages below.
# The same goes for other packages you might decide to use
# Remember that you only need to install each R package once in R (ever -- unless you change computers).
# All you need to do whenever you need to use the package again (after restarting the R session),
# is to use the library function to call the package.
# For example, type install.packages("knitr") in the console to install the knitr package. 
knitr::opts_chunk$set(echo = TRUE, fig.pos = 'h', fig.align = 'center')
knitr::opts_chunk$set(fig.cap = "",  fig.path = "Plot")
library(knitr)
library(dplyr)
library(arm)
library(pROC)
library(tidyverse)
library(MASS)
library(tigerstats)
library(leaps)
library(car)
library(rms)
require(caret)
require(e1071)
library(lme4) 
library(lattice)
library(broom)
library(boot)
library(ggplot2)
library(cobalt)
require(tidyverse)
require(rstanarm)
require(magrittr)
require(rstan)
require(MCMCpack)
library(abind)
library(matrixStats)
library(truncnorm)
library(mvtnorm)
library(MCMCpack)
library(mnormt)
```

* * *

1. Load data
essay: df of the raw data
grades: grade evaluated by different raters
char: characteristics of each work

```{r}
essay=read.csv("../DeYoreoKottas_supp/essay.csv",header=FALSE)
grades=essay[,1:5]
char=essay[,7:12]
names(char)=c("WL","SqW","PC","PS","PP","SL")
```

2. Extract features
p: Number of X (Covariates)
R: Number of Y (ordinal variables)
n: number of samples
x: df of covariate X
y: df of ordinal var Y
```{r}
p=0
R=3
d=p+R

n=nrow(essay)

y=matrix(0,n,R)

y[,1]=grades[,2]
y[,2]=grades[,3]
y[,3]=grades[,4]
```

3. Specify prior specification
N: Number of cluster in primary stick breaking process (DP)
R.x: Range of each covariates
mid.x: mid point of each covariates (max+min)/2
gamma1: cut off points of Y1
gamma2: cut off points of Y2
gamma3: cut off points of Y3
K: number of groups in ordinal data (3 levels: low med high)
R.z: range of z
```{r}
####prior specification####
N=40
Mon=500 # Final samples
B=50 # Burn-in
thin.int=10 # Thininig
source("../probitBayes.R")
sampled_z <- probitBayesImputation(data.frame(y), N, Mon, B, thin.int)
z1=sampled_z[,,1]
z2=sampled_z[,,2]
z3=sampled_z[,,3]
```

```{r}
sum(z1)
sum(z2)
sum(z3)
```

```{r}
head(z1)
```

* * *



