
---
title: "PMF of each ordinal variable"
output:
  pdf_document: default
  html_document:
    highlight: pygments
    theme: spacelab
---

```{r setup, echo =FALSE, include=FALSE}
# DO NOT ALTER CODE IN THIS CHUNK
# The very first time you use this R markdown file, you should install each of the packages below.
# The same goes for other packages you might decide to use
# Remember that you only need to install each R package once in R (ever -- unless you change computers).
# All you need to do whenever you need to use the package again (after restarting the R session),
# is to use the library function to call the package.
# For example, type install.packages("knitr") in the console to install the knitr package. 
knitr::opts_chunk$set(echo = TRUE, fig.pos = 'h', fig.align = 'center')
knitr::opts_chunk$set(fig.cap = "",  fig.path = "Plot")
library(knitr)
library(dplyr)
library(arm)
library(pROC)
library(tidyverse)
library(MASS)
library(tigerstats)
library(leaps)
library(car)
library(rms)
require(caret)
require(e1071)
library(lme4) 
library(lattice)
library(broom)
library(boot)
library(ggplot2)
library(cobalt)
require(tidyverse)
require(rstanarm)
require(magrittr)
require(rstan)
require(MCMCpack)
library(abind)
library(matrixStats)
library(truncnorm)
library(mvtnorm)
library(MCMCpack)
library(mnormt)
library(coda)
```

* * *
```{r, echo = FALSE}
load('../../Datasets/ordinalPUMS.Rdata')
n <- dim(df)[1]
```

### VEH: Vehicles available
1: No vehicles  
2: 1 vehicle  
3: 2 vehicles  
4: 3 vehicles  
5: 4 vehicles  
6: 5 vehicles  
7: 6 or more vehicles  

```{r, echo = FALSE}
# VEH: Vehicles
barplot(table(df$VEH)/n, xlab = 'level', ylab = 'pmf')
```

\newpage

### MV: When move into house
1: 12 months or less  
2: 13 to 23 months  
3: 2 to 4 years  
4: 5 to 9 years  
5: 10 to 19 years  
6: 20 to 29 years  
7: 30 years or more  

```{r, echo = FALSE}
# MV: When move into house
barplot(summary(df$MV)/n, xlab = 'level', ylab = 'pmf')
```

\newpage

### NP: Number of persons in this household
1: 1 person  
2: 2 persons  
3: 3 persons  
4: 4 persons  
5: 5 persons  
6: 6 persons  
7: 7 or more persons  

```{r, echo = FALSE}
# NP: Number of persons in this household
barplot(summary(df$NP)/n, xlab = 'level', ylab = 'pmf')
```

\newpage

### RMSP: Number of rooms
1 to 19 Number of rooms

```{r, echo = FALSE}
# RMSP: Number of rooms
barplot(summary(df$RMSP)/n, xlab = 'level', ylab = 'pmf')
```

\newpage

### ENG: Ability to speak english 
1: Not at all  
2: Not well  
3: Well  
4: Very well  
5: Speak only english  

```{r, echo = FALSE}
# ENG: Ability to speak english 
barplot(summary(df$ENG)/n, xlab = 'level', ylab = 'pmf')
```

\newpage

### MARHT: Number of times married
1: Never married  
2: One time  
3: Two times  
4: Three or more times  

```{r, echo = FALSE}
# MARHT: Number of times married 
barplot(summary(df$MARHT)/n, xlab = 'level', ylab = 'pmf')
```

\newpage

### SCHL: Educational attainment
1: No schooling complted  
2: Less than 12th grade  
3: High school, GED or college with no degree
4: Associate's degree  
5: Bachelor's degree  
6: Master's or other professional degree beyond bachelor's degree  
7: Doctorate degree  

```{r, echo = FALSE}
# SCHL: Educational attainment
barplot(summary(df$SCHL)/n, xlab = 'level', ylab = 'pmf')
```

\newpage

### RACNUM: Number of major race groups represented
1: One race group
2: Two or more race groups

```{r, echo = FALSE}
# RACNUM: Number of major race groups represented
barplot(summary(df$RACNUM)/n, xlab = 'level', ylab = 'pmf')
```

\newpage

### AGEP: Age
1: 15 to 19 years old  
2: 19 to 24 years old  
3: 25 to 29 years old  
4: 30 to 34 years old  
5: 35 to 39 years old  
6: 40 to 44 years old  
7: 45 to 49 years old  
8: 50 to 54 years old  
9: 55 to 59 years old  
10: 60 to 64 years old  
11: 65 to 69 years old  
12: 70 to 74 years old  
13: 75 to 79 years old  
14: 80 to 84 years old  
15: 85 to 89 years old  
16: 90 to 94 years old  
17: 94 to 99 years old  

```{r, echo = FALSE}
# AGEP: Age
barplot(summary(df$AGEP)/n,  xlab = 'level', ylab = 'pmf')
```

\newpage

### WKL: When last worked
1: Over 5 years ago or never worked  
2: 1-5 years ago  
3: Within the past 12 months  

```{r, echo = FALSE}
# WKL: When last worked
barplot(summary(df$WKL)/n,  xlab = 'level', ylab = 'pmf')
```

\newpage

### PINCP: Total person's income
1: Negative income  
2: None  
3: \$1 to \$20000  
4: \$20001 to \$40000   
5: \$40001 to \$60000   
6: \$60001 to \$80000   
7: \$80001 to \$100000   
8: \$100001 to \$120000   
9: \$120001 to \$140000   
10: \$140001 to \$160000   
11: \$160001 to \$180000   
12: \$180001 to \$200000   
13: More than \$200000  

```{r, echo = FALSE}
# PINCP: Total person's income
barplot(summary(df$PINCP)/n, xlab = 'level', ylab = 'pmf')
```

* * *



